Handbook streamFind: backend

### 2022-07-21 handbooks/Handbook_backend_streamFind.Rmd
LFU08 and new MS files and spiked chemicals

The streamFind package includes example files, which are located  in the folder `system.file(package = "streamFind", dir = "extdata")`. Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The files available are described in Table \@ref(tab:description-files).

@@

#list of file full paths
files <- list.files(r_path, full.names = TRUE)
```

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  file_name = basename(files),
  type = c(rep("MS/MS", 9)),
  polarity = c(rep("positive", 9)),
  description = c(
    rep("Basic centroided MS data spiked with chemicals.", 3),
    rep("Blank centroided MS data spiked with internal standards.", 3),
    rep("Wastewater secondary effluent centroided MS data spiked with internal standards.", 3)
  )
)

@@

`r # TODO add a list of spiked chemical in the package` List the chemicals spiked in the MS data files.

@@

```{r make_targets}
#case 1
mz1 <- c(247.1651, 239.0628)
rt1 <- c(839, 937)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = 20, sec = 60, id = id1)

@@

#case 2
mz2 <- data.frame(mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652))
rt2 <- data.frame(rtmin = c(809, 907), rtmax = c(869, 967))
targets2 <- makeTargets(mz = mz2, rt = rt2)

@@

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(247.1651, 239.0628),
  rt = c(839, 937)
)
targets3 <- makeTargets(mz = mz3, ppm = 20, sec = 60)

@@

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652),
  rtmin = c(809, 907), rtmax = c(869, 967)
)
targets4 <- makeTargets(mz = mz4)

@@

```{r newStreamProject_basic_workflow, message=FALSE, results='hide'}
exS2 <- newStreamProject(files = files[4:9], path = getwd(), title = "Project Example")
```

@@

```{r assign_replicates}
#getter for replicate names in msData
replicates(exS2)

#setter for replicate names in msData
replicates(exS2) <- c(rep("blank", 3),rep("influent", 3))  
replicates(exS2)

#setter for replicate name in msAnalysis
exA2 <- getAnalyses(exS2, 4)
replicates(exA2) <- "wastewater"

#getter for replicate name in msAnalysis
replicates(exA2)
```

@@

```{r assign_blanks}
#getter for blank replicates in each sample in a msData object
blanks(exS2)

#setter for the blank analysis replicate of msData object
blanks(exS2) <- rep("blank", 6)
blanks(exS2)

#getter for msAnalysis
blanks(exA2)
```

@@

```{r other_S4_methods}
#getter for analysis names in msdata
analyses(exS2)

#getter for analysis name in msAnalysis
analyses(exA2)

#getter for polarity of analyses in a msData
polarities(exS2)

#getter for polarity of an msAnalysis
polarities(exA2)

#subsetting an msData with `[`-method for the blank samples
exS2[1:3]
```

@@

exS2_pp <- exS2

@@

# add the settings to all analyses in msData
exS2_pp <- addParameters(exS2_pp, settings = settings_pp, where = "analyses")

# get the parameters in the third analysis of the msData object
getParameters(exS2_pp, where = "analyses", analyses = 3)

# add the settings to the msAnalysis object
exA2 <- addParameters(exA2, settings = settings_pp)

# get the parameters of the msAnalysis object
getParameters(exA2, call = "peakPicking")
```

@@

#peak picking call using the stored parameters
exS2_pp <- peakPicking(exS2_pp)

# peaks added for each analysis, as shown by column peaks of show method (below)
exS2_pp

#alternatively, parameters can be added during the function call
exS2_pp_2 <- peakPicking(exS2, settings = settings_pp)

# parameters are added to each analysis after running peak picking
summary(getParameters(exS2_pp_2, where = "analyses", analyses = 1))

exS2_pp_2

# running peak picking for an msAnalysis object (settings were already added above)
exA2 <- peakPicking(exA2)

# msAnalysis with peaks
exA2
```

@@

exS2_pp_3 <- exS2

@@

exS2_pp_3 <- addParameters(exS2_pp_3, settings = settings_pp, where = "analyses")

# adding different settings to analysis 4 to 6
exS2_pp_3 <- addParameters(exS2_pp_3, settings = settings_pp_2, where = "analyses", analyses = 4:6)

# the algorithm of the influent samples is different
sapply(analyses(exS2_pp_3), function(x) getAlgorithm(getParameters(exS2_pp_3, where = "analyses", analyses = x)[[1]]))

#running peak picking
exS2_pp_3 <- peakPicking(exS2_pp_3)

# as the peak parameters with "openms" as algorithm were less stringent, the influent samples have more peaks then when "xcms3" is used 
exS2_pp_3

@@

```{r inspecting_peaks}
#access to peaks in msData based on built targets as used for extracting EICs
peaks(exS2_pp[c(1, 4)], mz = mz1[1], rt = rt1[1], ppm = 10, sec = 30)

#access to peak in msAnalysis
peaks(exA2, mz = targets4)

#plotting peaks in msData
plotPeaks(exS2_pp, mz = targets4[2, ], interactive = TRUE, colorBy = "analyses")

#plotting peaks from msAnalysis
plotPeaks(exA2, mz = targets4, interactive = FALSE)

#map plot for peak time and m/z dimensions for msAnalysis
mapPeaks(exA2, mz = targets4[1, ], xlim = 30, ylim = 0.001)

##map plot for peak time and m/z dimensions for msData
mapPeaks(exS2_pp, mz = targets4[2, ], colorBy = "replicates", xlim = 30, ylim = 0.001)
```

@@

# add processing parameter for peakGrouping to features in the msData
exS2_pp <- addParameters(exS2_pp, settings = settings_pg, where = "features")

# get the processing parameters applied to features
getParameters(exS2_pp, where = "features")
```

@@

```{r processing_grouping, message=FALSE}
#using the added parameters
exS2_pg <- peakGrouping(exS2_pp)

#adding the parameters as arguments
exS2_pg_2 <- peakGrouping(exS2_pp, settings = settings_pg)
```

@@

```{r msFeatures_class}
#accessing directly the msFeatures slot, not recommended
str(exS2_pg@features, max.level = 2)
```

@@

```{r inspecting_features}
#getter for feature intensities and intensity deviations
features(exS2_pg, mz = targets4)

#getter for all the feature metadata by setting complete to TRUE
t(features(exS2_pg_2, mz = targets4[1, ], complete = TRUE))

#getter for features with intensities for each sample by setting average to FALSE
features(exS2_pg_2, mz = targets4[1, ], average = FALSE)

#plotting features
plotFeatures(exS2_pg, mz = targets4, colorBy = "targets", interactive = FALSE)

#plotting features for replicates and interactive
plotFeatures(exS2_pg, mz = targets4, colorBy = "replicates", interactive = TRUE)

#plot the individual peaks in features and for each analysis shows the time deviation.
plotFeaturePeaks(exS2_pg, mz = targets4)
```

@@

```{r plot_alignment}
plotAlignment(exS2_pg)
```

@@

```{r recursive_integration_procesing, results='hide', message=FALSE}
#filling peaks for msData
exS2_pg_fl <- peakFilling(exS2_pg, settings = settings_fl)
```

@@

```{r recursive_integration_inspection}
# inspecting the filled peaks for a given mz and rt target
plotFeaturePeaks(exS2_pg_fl, mz = 441.1670, rt = 916, ppm = 20, sec = 60)
```

@@

exS2_pg_fl_an <- peakAnnotation(exS2_pg_fl, settings = settings_an)
```

@@

```{r annotation_results}
#isotopic and adduct features annotated with a given target mz and rt
t(streamFind::annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE))

#plotting the annotation of each feature in a given target
plotAnnotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
```

@@

```{r filter_subsetting}
#sub-setting the msData class on analyses (the 3 first)
exS2_pg[1:3, ]

#sub-setting the msData class on analys2s and features (the 3 last analyses and features annotated with a target)
target_s <- streamFind::annotation(exS2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
exS2_pg_fl_an[4:6, target_s$id]

#sub-seeting the msData class on analyses ([i, ]) and features ([, j]) by index
exS2_pg_fl_an[1:2, 1:5]

#sub-setting an msAnalysis class on the first 5 peaks
exA2[1:5]
```

@@

exS2_pg_fl_an_filFeat <- addParameters(exS2_pg_fl_an, where = "features", settings = settings_filFeat)

exS2_pg_fl_an_filFeat <- filterFeatures(exS2_pg_fl_an_filFeat)

#number of features not filtered
nrow(features(exS2_pg_fl_an, filtered = FALSE))

#number of features not filtered after applying the filters
nrow(features(exS2_pg_fl_an_filFeat, filtered = FALSE))

#remove permanently all filtered features from the msData object
exS2_pg_fl_an_filFeat_rem <- removeFilteredFeatures(exS2_pg_fl_an_filFeat, which =  "all")

exS2_pg_fl_an_filFeat_rem
```

@@

# exS2_pg <- calculateSNR(exS2_pg, targetsID = features(exS2_pg)[["id"]][1:5])
# 
# features(exS2_pg, complete = TRUE)[1:5, ]

@@

# object <- exS2_pg_fl
# settings <- annotationSettingsDefaultCAMERA()
# settings <- annotationSettingsDefaultRAMClustR()
# object <- addParameters(object, where = "features", settings)
# 
# 
# 
# findFGroup(comp, "M239_R936_497")
# as.data.frame(comp[702, ])
# 
# mapPeaks(object, mz = data.table(mzmin = 208, mzmax = 220, rtmin = 635, rtmax = 645))
# 
# plotXICs(object, analyses = 1:2, mz = 262.0450, ppm = 40, rt = 936, sec = 120)
# 
# View(features(object, mz = 441.1670, ppm = 20, rt = 916, sec = 10, complete = TRUE))
# View(features(object, complete = TRUE)[monoiso %in% "m441.168_d0.7_r916_t6_f3225", ])

>>>>

The streamFind package includes example files, which are located  in the folder `system.file(package = "streamFind", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, _m/z_ for MS1 data between 200 and 450 Da and _m/z_ for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, _m/z_ for MS1 data between 200 and 300 Da and _m/z_ for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the backend.

@@

#list of file full paths
files <- list.files(r_path, pattern = ".mzML", full.names = TRUE)
```

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  file_name = basename(files),
  device = c(rep("Agilent Q-TOF", 24)),
  type = c(rep("centroided", 3), rep("profile", 3), rep("profile", 18)),
  mode = c(rep("MS/MS", 24)),
  polarity = c(rep("positive", 6),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3)),
  description = c(
    rep("Basic centroided MS data spiked with chemicals.", 3),
    rep("Basic profile MS data spiked with chemicals.", 3),
    rep("Blank MS data spiked with internal standards.", 6),
    rep("Wastewater secondary effluent MS data spiked with internal standards.", 6),
    rep("Wastewater secondary effluent treated with ozonated strong water MS data spiked with internal standards.", 6)
  )
)

@@

The chemicals spiked in the files are described below in Table \@ref(tab:show-table-chemicals). The internal standards (IS) are spiked to all Agilent Q-TOF files. The chemical standards (S) were only spiked for the first 6  Agilent Q-TOF files not to the blanks nor wastewater analysis files. Note that not all IS and S are visible in the profile Agilent Q-TOF files as a narrower trimming was applied.

```{r load-table-chemicals}
db <- paste0(r_path, "/spiked_chemicals_hrms.csv")
db <- data.table::fread(db)
```

```{r show-table-chemicals, echo=FALSE}
db$ionization <- "positive"
db[, `:=`("mz_pos" = neutralMass + 1.0073, "mz_neg" = neutralMass - 1.0073 )]
db[grepl("neg", db$comment), ionization := "both"]
data.table::setnames(db, "mixIUTA", "tag")
db[tag %in% 1, tag := "S"]
knitr::kable(db[, .(name, PubChem, formula, neutralMass, rt, ionization, tag)], caption = "Chemicals spiked.") %>%
kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "condensed", "responsive"), fixed_thead = TRUE)
```

@@

```{r make_targets}
#case 1
mz1 <- c(247.1651, 239.0628)
rt1 <- c(1075, 1157)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = 20, sec = 60, id = id1)

@@

#case 2
mz2 <- data.frame(mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652))
rt2 <- data.frame(rtmin = c(1045, 1127), rtmax = c(1105, 1187))
targets2 <- makeTargets(mz = mz2, rt = rt2)

@@

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(247.1651, 239.0628),
  rt = c(1075, 1157)
)
targets3 <- makeTargets(mz = mz3, ppm = 20, sec = 60)

@@

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652),
  rtmin = c(1045, 1127), rtmax = c(1105, 1187)
)
targets4 <- makeTargets(mz = mz4)

@@

```{r newStreamProject_basic_workflow, message=FALSE, results='hide'}
sp2 <- newStreamProject(files = files[c(10:12, 16:18)], path = getwd(), title = "Project Example")
```

@@

```{r assign_replicates}
#getter for replicate names in msData
replicates(sp2)

#setter for replicate names in msData
replicates(sp2) <- c(rep("blank", 3),rep("influent", 3))  
replicates(sp2)

#setter for replicate name in msAnalysis
a2 <- getAnalyses(sp2, 4)
replicates(a2) <- "wastewater"

#getter for replicate name in msAnalysis
replicates(a2)
```

@@

```{r assign_blanks}
#getter for blank replicates in each sample in a msData object
blanks(sp2)

#setter for the blank analysis replicate of msData object
blanks(sp2) <- rep("blank", 6)
blanks(sp2)

#getter for msAnalysis
blanks(a2)
```

@@

```{r other_S4_methods}
#getter for analysis names in msdata
analyses(sp2)

#getter for analysis name in msAnalysis
analyses(a2)

#getter for polarity of analyses in a msData
polarities(sp2)

#getter for polarity of an msAnalysis
polarities(a2)

#subsetting an msData with `[`-method for the blank samples
sp2[1:3]
```

@@

sp2_pp <- sp2

@@

# add the settings to all analyses in msData
sp2_pp <- addParameters(sp2_pp, settings = settings_pp, where = "analyses")

# get the parameters in the third analysis of the msData object
getParameters(sp2_pp, where = "analyses", analyses = 3)

# add the settings to the msAnalysis object
a2 <- addParameters(a2, settings = settings_pp)

# get the parameters of the msAnalysis object
getParameters(a2, call = "peakPicking")
```

@@

#peak picking call using the stored parameters
sp2_pp <- peakPicking(sp2_pp)

# peaks added for each analysis, as shown by column peaks of show method (below)
sp2_pp

#alternatively, parameters can be added during the function call
sp2_pp_2 <- peakPicking(sp2, settings = settings_pp)

# parameters are added to each analysis after running peak picking
summary(getParameters(sp2_pp_2, where = "analyses", analyses = 1))

sp2_pp_2

# running peak picking for an msAnalysis object (settings were already added above)
a2 <- peakPicking(a2)

# msAnalysis with peaks
a2
```

@@

sp2_pp_3 <- sp2

@@

sp2_pp_3 <- addParameters(sp2_pp_3, settings = settings_pp, where = "analyses")

# adding different settings to analysis 4 to 6
sp2_pp_3 <- addParameters(sp2_pp_3, settings = settings_pp_2, where = "analyses", analyses = 4:6)

# the algorithm of the influent samples is different
sapply(analyses(sp2_pp_3), function(x) getAlgorithm(getParameters(sp2_pp_3, where = "analyses", analyses = x)[[1]]))

#running peak picking
sp2_pp_3 <- peakPicking(sp2_pp_3)

# as the peak parameters with "openms" as algorithm were less stringent, the influent samples have more peaks then when "xcms3" is used 
sp2_pp_3

@@

```{r inspecting_peaks}
#access to peaks in msData based on built targets as used for extracting EICs
peaks(sp2_pp[c(1, 4)], mz = mz1[1], rt = rt1[1], ppm = 10, sec = 30)

#access to peak in msAnalysis
peaks(a2, mz = targets4)

#plotting peaks in msData
plotPeaks(sp2_pp, mz = targets4[2, ], interactive = TRUE, colorBy = "analyses")

#plotting peaks from msAnalysis
plotPeaks(a2, mz = targets4, interactive = FALSE)

#map plot for peak time and m/z dimensions for msAnalysis
mapPeaks(a2, mz = targets4[1, ], xlim = 30, ylim = 0.001)

##map plot for peak time and m/z dimensions for msData
mapPeaks(sp2_pp, mz = targets4[2, ], colorBy = "replicates", xlim = 30, ylim = 0.001)
```

@@

# add processing parameter for peakGrouping to features in the msData
sp2_pp <- addParameters(sp2_pp, settings = settings_pg, where = "features")

# get the processing parameters applied to features
getParameters(sp2_pp, where = "features")
```

@@

```{r processing_grouping, message=FALSE}
#using the added parameters
sp2_pg <- peakGrouping(sp2_pp)

#adding the parameters as arguments
sp2_pg_2 <- peakGrouping(sp2_pp, settings = settings_pg)
```

@@

```{r msFeatures_class}
#accessing directly the msFeatures slot, not recommended
str(sp2_pg@features, max.level = 2)
```

@@

```{r inspecting_features}
#getter for feature intensities and intensity deviations
features(sp2_pg, mz = targets4)

#getter for all the feature metadata by setting complete to TRUE
t(features(sp2_pg_2, mz = targets4[1, ], complete = TRUE))

#getter for features with intensities for each sample by setting average to FALSE
features(sp2_pg_2, mz = targets4[1, ], average = FALSE)

#plotting features
plotFeatures(sp2_pg, mz = targets4, colorBy = "targets", interactive = FALSE)

#plotting features for replicates and interactive
plotFeatures(sp2_pg, mz = targets4, colorBy = "replicates", interactive = TRUE)

#plot the individual peaks in features and for each analysis shows the time deviation.
plotFeaturePeaks(sp2_pg, mz = targets4)
```

@@

```{r plot_alignment}
plotAlignment(sp2_pg)
```

@@

```{r recursive_integration_procesing, results='hide', message=FALSE}
#filling peaks for msData
sp2_pg_fl <- peakFilling(sp2_pg, settings = settings_fl)
```

@@

```{r recursive_integration_inspection}
# inspecting the filled peaks for a given mz and rt target
plotFeaturePeaks(sp2_pg_fl, mz = 441.1670, rt = 916, ppm = 20, sec = 60)
```

@@

sp2_pg_fl_an <- peakAnnotation(sp2_pg_fl, settings = settings_an)
```

@@

```{r annotation_results}
#isotopic and adduct features annotated with a given target mz and rt
t(streamFind::annotation(sp2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE))

#plotting the annotation of each feature in a given target
plotAnnotation(sp2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
```

@@

```{r filter_subsetting}
#sub-setting the msData class on analyses (the 3 first)
sp2_pg[1:3, ]

#sub-setting the msData class on analys2s and features (the 3 last analyses and features annotated with a target)
target_s <- streamFind::annotation(sp2_pg_fl_an, mz = 441.1670, ppm = 10, rt = 916, sec = 5, all = FALSE)
sp2_pg_fl_an[4:6, target_s$id]

#sub-seeting the msData class on analyses ([i, ]) and features ([, j]) by index
sp2_pg_fl_an[1:2, 1:5]

#sub-setting an msAnalysis class on the first 5 peaks
a2[1:5]
```

@@

sp2_pg_fl_an_filFeat <- addParameters(sp2_pg_fl_an, where = "features", settings = settings_filFeat)

sp2_pg_fl_an_filFeat <- filterFeatures(sp2_pg_fl_an_filFeat)

#number of features not filtered
nrow(features(sp2_pg_fl_an, filtered = FALSE))

#number of features not filtered after applying the filters
nrow(features(sp2_pg_fl_an_filFeat, filtered = FALSE))

#remove permanently all filtered features from the msData object
sp2_pg_fl_an_filFeat_rem <- removeFilteredFeatures(sp2_pg_fl_an_filFeat, which =  "all")

sp2_pg_fl_an_filFeat_rem
```

@@

# sp2_pg <- calculateSNR(sp2_pg, targetsID = features(sp2_pg)[["id"]][1:5])
# 
# features(sp2_pg, complete = TRUE)[1:5, ]

@@

# object <- sp2_pg_fl
# settings <- annotationSettingsDefaultCAMERA()
# settings <- annotationSettingsDefaultRAMClustR()
# object <- addParameters(object, where = "features", settings)
# 
# 
# 
# findFGroup(comp, "M239_R936_497")
# as.data.frame(comp[702, ])
# 
# mapPeaks(object, mz = data.table(mzmin = 208, mzmax = 220, rtmin = 635, rtmax = 645))
# 
# plotXICs(object, analyses = 1:2, mz = 262.0450, ppm = 40, rt = 936, sec = 120)
# 
# View(features(object, mz = 441.1670, ppm = 20, rt = 916, sec = 10, complete = TRUE))
# View(features(object, complete = TRUE)[monoiso %in% "m441.168_d0.7_r916_t6_f3225", ])
